---
title: "ECON 470 Homework 4"
subtitle: "Submission 1, Spring 2026"
author: "Sophie Kim"
format:
  pdf:
    output-file: "kim-s-hwk4-1"
    output-ext:  "pdf"
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
---

```{r}
#| include: false

# Preliminaries -----------------------------------------------------------
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, haven)
```


## Problem 1


```{r}
#| echo: false
#| warning: false

final.hcris <- read_tsv("../data/output/HCRIS_Data.txt")

dupes_by_year <- final.hcris %>%
  mutate(year = year(fy_end)) %>%
  filter(year >= 2009, year <= 2019) %>%
  filter(source == "total for year") %>%
  group_by(year) %>%
  summarise(n_hospitals = n_distinct(provider_number))

ggplot(dupes_by_year, aes(x = year, y = n_hospitals)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Hospitals filing more than one report per year",
    x = "Year",
    y = "Number of hospitals"
  ) +
  theme_minimal()
```

\newpage
## Problem 2

```{r}
#| echo: false
#| warning: false

final.hcris %>%
  mutate(year = year(fy_end)) %>%
  filter(year >= 2009, year <= 2019) %>%
  summarise(unique_hospitals = n_distinct(provider_number))
```

\newpage
## Problem 3

```{r}
#| echo: false
#| warning: false

final.hcris %>%
  mutate(year = year(fy_end)) %>%
  filter(year >= 2009, year <= 2019) %>%
  ggplot(aes(x = factor(year), y = tot_charges)) +
  geom_violin() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of total charges by year",
    x = "Year",
    y = "Total charges"
  ) +
  theme_minimal()
```

\newpage
## Problem 4

```{r}
#| echo: false
#| warning: false

final.hcris %>%
  mutate(
    year = year(fy_end),
    discount_factor = 1 - tot_discounts / tot_charges,
    price_num = (ip_charges + icu_charges + ancillary_charges) * discount_factor - tot_mcare_payment,
    price_denom = tot_discharges - mcare_discharges,
    price = price_num / price_denom
  ) %>%
  filter(year >= 2009, year <= 2019) %>%
  filter(price > 0) %>%
  group_by(year) %>%
  mutate(
    p05 = quantile(price, 0.05, na.rm = TRUE),
    p95 = quantile(price, 0.95, na.rm = TRUE),
    price = pmin(pmax(price, p05), p95)
  ) %>%
  ggplot(aes(x = factor(year), y = price)) +
  geom_violin() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of estimated hospital prices by year",
    x = "Year",
    y = "Estimated price"
  ) +
  theme_minimal()
```

\newpage
## Problem 5

```{r}
#| echo: false
#| warning: false

final.hcris %>%
  mutate(year = year(fy_end)) %>%
  filter(year >= 2012, year <= 2019) %>%
  group_by(year) %>%
  summarise(
    total = n(),
    penalized = sum(hrrp_payment > 0, na.rm = TRUE),
    share_penalized = penalized / total
  ) %>%
  ggplot(aes(x = year, y = share_penalized)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Share of hospitals penalized under HRRP/VBP",
    x = "Year",
    y = "Share penalized"
  ) +
  theme_minimal()
```

\newpage
## Problem 6

```{r}
#| include: false

#Calculate price for each hospital-year
hcris.price <- final.hcris %>%
  mutate(
    year = year(fy_end),
    discount_factor = 1 - tot_discounts / tot_charges,
    price_num = (ip_charges + icu_charges + ancillary_charges) * discount_factor - tot_mcare_payment,
    price_denom = tot_discharges - mcare_discharges,
    price = price_num / price_denom
  ) %>%
  filter(price > 0) %>%
  group_by(year) %>%
  mutate(
    p05 = quantile(price, 0.05, na.rm = TRUE),
    p95 = quantile(price, 0.95, na.rm = TRUE),
    price = pmin(pmax(price, p05), p95)
  ) %>%
  ungroup() %>%
  select(provider_number, year, price, beds, mcare_discharges, mcaid_discharges, hrrp_payment, hvbp_payment)

price.change <- hcris.price %>%
  filter(year %in% c(2011, 2014)) %>%
  select(provider_number, year, price) %>%
  pivot_wider(names_from = year, values_from = price, names_prefix = "price_") %>%
  mutate(price_change = price_2014 - price_2011)

penalty.2012 <- hcris.price %>%
  filter(year == 2012) %>%
  mutate(net_penalty = hrrp_payment + hvbp_payment) %>%
  select(provider_number, net_penalty)

pre.2012 <- hcris.price %>%
  filter(year >= 2009, year <= 2011) %>%
  group_by(provider_number) %>%
  summarise(
    avg_beds = mean(beds, na.rm = TRUE),
    avg_mcare = mean(mcare_discharges, na.rm = TRUE),
    avg_mcaid = mean(mcaid_discharges, na.rm = TRUE)
  )

hcris.final <- price.change %>%
  inner_join(penalty.2012, by = "provider_number") %>%
  inner_join(pre.2012, by = "provider_number") %>%
  filter(!is.na(price_change))
```

```{r}
#| include: false

ols1 <- lm(price_change ~ net_penalty, data = hcris.final)
ols2 <- lm(price_change ~ net_penalty + avg_beds, data = hcris.final)
ols3 <- lm(price_change ~ net_penalty + avg_beds + avg_mcaid, data = hcris.final)

summary(ols1)
summary(ols2)
summary(ols3)
```

```{r}
#| echo: false
#| warning: false

options(knitr.kable.NA = 0)
library(knitr)
library(kableExtra)

# Helper to format p-values
fmt_p <- function(p) ifelse(is.na(p), NA, sprintf("%.3f", p))

# Extract p-values
p1 <- coef(summary(ols1))[, "Pr(>|t|)"]
p2 <- coef(summary(ols2))[, "Pr(>|t|)"]
p3 <- coef(summary(ols3))[, "Pr(>|t|)"]

data.frame(
  Statistic = c(
    "Intercept", "p-value",
    "Net Penalty", "p-value",
    "Mean Beds", "p-value",
    "Mean Medicaid", "p-value",
    "# Obs", "R2"
  ),
  `OLS 1` = c(
    round(coef(ols1)["(Intercept)"], 3), fmt_p(p1["(Intercept)"]),
    round(coef(ols1)["net_penalty"], 3), fmt_p(p1["net_penalty"]),
    NA, NA,
    NA, NA,
    nobs(ols1), round(summary(ols1)$r.squared, 3)
  ),
  `OLS 2` = c(
    round(coef(ols2)["(Intercept)"], 3), fmt_p(p2["(Intercept)"]),
    round(coef(ols2)["net_penalty"], 3), fmt_p(p2["net_penalty"]),
    round(coef(ols2)["avg_beds"],    3), fmt_p(p2["avg_beds"]),
    NA, NA,
    nobs(ols2), round(summary(ols2)$r.squared, 3)
  ),
  `OLS 3` = c(
    round(coef(ols3)["(Intercept)"], 3), fmt_p(p3["(Intercept)"]),
    round(coef(ols3)["net_penalty"], 3), fmt_p(p3["net_penalty"]),
    round(coef(ols3)["avg_beds"],    3), fmt_p(p3["avg_beds"]),
    round(coef(ols3)["avg_mcaid"],   3), fmt_p(p3["avg_mcaid"]),
    nobs(ols3), round(summary(ols3)$r.squared, 3)
  ),
  check.names = FALSE
) %>%
  kable(
    col.names = c("Statistic", "OLS 1", "OLS 2", "OLS 3"),
    digits    = 3,
    booktabs  = TRUE,
    caption   = "OLS Results of Penalty Status from Price Change"
  ) %>%
  kable_styling(latex_options = c("hold_position", "striped")) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE) %>%
  # Style p-value rows differently (rows 2, 4, 6, 8)
  row_spec(c(2, 4, 6, 8), italic = TRUE, color = "gray") %>%
  pack_rows("Coefficients", 1, 8) %>%
  pack_rows("Model Fit",    9, 10)
```

\newpage

## Problem 7

```{r}
#| echo: false
#| warning: false

ggplot(hcris.final, aes(x = avg_mcare, y = net_penalty)) +
  geom_point(alpha = 0.3) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Net penalty vs pre-2012 Medicare discharges",
    x = "Average Medicare discharges (2009-2011)",
    y = "Net penalty (2012)"
  ) +
  theme_minimal()
```

\newpage
## Problem 8

```{r}
#| include: false

library(fixest)

# First stage: net_penalty ~ avg_mcare (instrument)
fs1 <- feols(net_penalty ~ avg_mcare, data = hcris.final)
fs2 <- feols(net_penalty ~ avg_mcare + avg_beds, data = hcris.final)
fs3 <- feols(net_penalty ~ avg_mcare + avg_beds + avg_mcaid, data = hcris.final)

# Reduced form: price_change ~ avg_mcare (instrument)
rf1 <- feols(price_change ~ avg_mcare, data = hcris.final)
rf2 <- feols(price_change ~ avg_mcare + avg_beds, data = hcris.final)
rf3 <- feols(price_change ~ avg_mcare + avg_beds + avg_mcaid, data = hcris.final)

cat("=== First Stage ===\n")
summary(fs1)
summary(fs2)
summary(fs3)

cat("=== Reduced Form ===\n")
summary(rf1)
summary(rf2)
summary(rf3)
```

```{r}
#| echo: false
#| warning: false

options(knitr.kable.NA = 0)
library(knitr)
library(kableExtra)

fmt_p <- function(p) ifelse(is.na(p), NA, sprintf("%.3f", p))

get_coef <- function(model, var) {
  ct <- coef(model)
  if (var %in% names(ct)) round(ct[var], 3) else NA
}
get_p <- function(model, var) {
  pt <- coeftable(model)[, "Pr(>|t|)"]
  if (var %in% names(pt)) fmt_p(pt[var]) else NA
}

data.frame(
  Statistic = c("Intercept", "p-value", "Avg Medicare", "p-value", "Mean Beds", "p-value", "Mean Medicaid", "p-value", "# Obs", "R2"),
  `FS 1` = c(get_coef(fs1, "(Intercept)"), get_p(fs1, "(Intercept)"), get_coef(fs1, "avg_mcare"), get_p(fs1, "avg_mcare"), NA, NA, NA, NA, nobs(fs1), round(r2(fs1, "r2"), 3)),
  `FS 2` = c(get_coef(fs2, "(Intercept)"), get_p(fs2, "(Intercept)"), get_coef(fs2, "avg_mcare"), get_p(fs2, "avg_mcare"), get_coef(fs2, "avg_beds"), get_p(fs2, "avg_beds"), NA, NA, nobs(fs2), round(r2(fs2, "r2"), 3)),
  `FS 3` = c(get_coef(fs3, "(Intercept)"), get_p(fs3, "(Intercept)"), get_coef(fs3, "avg_mcare"), get_p(fs3, "avg_mcare"), get_coef(fs3, "avg_beds"), get_p(fs3, "avg_beds"), get_coef(fs3, "avg_mcaid"), get_p(fs3, "avg_mcaid"), nobs(fs3), round(r2(fs3, "r2"), 3)),
  `RF 1` = c(get_coef(rf1, "(Intercept)"), get_p(rf1, "(Intercept)"), get_coef(rf1, "avg_mcare"), get_p(rf1, "avg_mcare"), NA, NA, NA, NA, nobs(rf1), round(r2(rf1, "r2"), 3)),
  `RF 2` = c(get_coef(rf2, "(Intercept)"), get_p(rf2, "(Intercept)"), get_coef(rf2, "avg_mcare"), get_p(rf2, "avg_mcare"), get_coef(rf2, "avg_beds"), get_p(rf2, "avg_beds"), NA, NA, nobs(rf2), round(r2(rf2, "r2"), 3)),
  `RF 3` = c(get_coef(rf3, "(Intercept)"), get_p(rf3, "(Intercept)"), get_coef(rf3, "avg_mcare"), get_p(rf3, "avg_mcare"), get_coef(rf3, "avg_beds"), get_p(rf3, "avg_beds"), get_coef(rf3, "avg_mcaid"), get_p(rf3, "avg_mcaid"), nobs(rf3), round(r2(rf3, "r2"), 3)),
  check.names = FALSE
) %>%
  kable(
    col.names = c("Statistic", "FS 1", "FS 2", "FS 3", "RF 1", "RF 2", "RF 3"),
    digits  = 3,
    booktabs = TRUE,
    caption = "First Stage and Reduced Form Results"
  ) %>%
  kable_styling(latex_options = c("hold_position", "striped")) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE) %>%
  row_spec(c(2, 4, 6, 8), italic = TRUE, color = "gray") %>%
  add_header_above(c(" " = 1, "First Stage" = 3, "Reduced Form" = 3)) %>%
  pack_rows("Coefficients", 1, 8) %>%
  pack_rows("Model Fit", 9, 10)
```

\newpage

## Problem 9

```{r}
#| include: false

# IV estimates: price_change ~ net_penalty, instrumented by avg_mcare
iv1 <- feols(price_change ~ 1 | net_penalty ~ avg_mcare, data = hcris.final)
iv2 <- feols(price_change ~ avg_beds | net_penalty ~ avg_mcare, data = hcris.final)
iv3 <- feols(price_change ~ avg_beds + avg_mcaid | net_penalty ~ avg_mcare, data = hcris.final)

summary(iv1)
summary(iv2)
summary(iv3)
```

```{r}
#| echo: false
#| warning: false

options(knitr.kable.NA = 0)
library(knitr)
library(kableExtra)

fmt_p <- function(p) ifelse(is.na(p), NA, sprintf("%.3f", p))

get_coef <- function(model, var) {
  ct <- coef(model)
  if (var %in% names(ct)) round(ct[var], 3) else NA
}
get_p <- function(model, var) {
  pt <- coeftable(model)[, "Pr(>|t|)"]
  if (var %in% names(pt)) fmt_p(pt[var]) else NA
}

data.frame(
  Statistic = c("Intercept", "p-value", "Net Penalty", "p-value", "Mean Beds", "p-value", "Mean Medicaid", "p-value", "# Obs", "R2"),
  `IV 1` = c(get_coef(iv1, "(Intercept)"), get_p(iv1, "(Intercept)"), get_coef(iv1, "fit_net_penalty"), get_p(iv1, "fit_net_penalty"), NA, NA, NA, NA, nobs(iv1), round(r2(iv1, "r2"), 3)),
  `IV 2` = c(get_coef(iv2, "(Intercept)"), get_p(iv2, "(Intercept)"), get_coef(iv2, "fit_net_penalty"), get_p(iv2, "fit_net_penalty"), get_coef(iv2, "avg_beds"), get_p(iv2, "avg_beds"), NA, NA, nobs(iv2), round(r2(iv2, "r2"), 3)),
  `IV 3` = c(get_coef(iv3, "(Intercept)"), get_p(iv3, "(Intercept)"), get_coef(iv3, "fit_net_penalty"), get_p(iv3, "fit_net_penalty"), get_coef(iv3, "avg_beds"), get_p(iv3, "avg_beds"), get_coef(iv3, "avg_mcaid"), get_p(iv3, "avg_mcaid"), nobs(iv3), round(r2(iv3, "r2"), 3)),
  check.names = FALSE
) %>%
  kable(
    col.names = c("Statistic", "IV 1", "IV 2", "IV 3"),
    digits  = 3,
    booktabs = TRUE,
    caption = "IV Results of Penalty Status on Price Change"
  ) %>%
  kable_styling(latex_options = c("hold_position", "striped")) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE) %>%
  row_spec(c(2, 4, 6, 8), italic = TRUE, color = "gray") %>%
  pack_rows("Coefficients", 1, 8) %>%
  pack_rows("Model Fit", 9, 10)
```

\newpage

## Problem 10

Briefly explain the “Local” ATE in the context of your estimates. How might a local effect differ from an overall ATE in this setting?

Local ATE would be the average treatment effect that are only for hospitals that were penalized because of their Medicare share. This would capture the effect of penalty status on price changes only for hospitals that were penalized from Medicare patient share, making it a narrower group than all penalized hospitals. The local effect could be larger than ATE if the hospitals within the local group is more sensitive to Medicare reimbursement changes. 


